---
title: "Soybean Yield Gap Analysis Clean Code"
author: "Sam Wallace"
date: "2025-01-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown includes "clean", as in more direct code of the R code files in the Soybean_yga/Code folder for more centralized and cleaner use in analysis. R markdown is created by Sam Wallace

```{r Libraries,echo=F}
library(tidyverse)
library(emmeans)
library(lme4)
library(modelr)
library(ggpmisc)
library(lmerTest)
library(glmmTMB)
library(DHARMa)
library(shades)
library(meta)
library(metafor)
library(here)
library(ggpubr)
library(performance)
library(janitor)
library(fitdistrplus)

options(contrasts = c("contr.sum", "contr.poly")) #setting contrasts
```


```{r Read in PRISM weather data}
#### PRISM weather data was extracted from python based API provided by Ben Eck from NRCC. The files here are raw data from there and are cleaned in this chunk.

PrismRaw2Clean <- function(csvpath, location){
  
  # Reads in raw file based on the file path including the file name and .csv. The here function is locating the r project location, so anything in the path will come after the project location in your file path.
  raw_prism = read_csv(here(csvpath))
  colnames(raw_prism) = c("date", "maxt", "mint", "prcp")
  raw_prism$date <- as.Date(raw_prism$date, format = "%m/%d/%Y")
  
  # This converts the english units from PRISM into metric and also finds the day of year for each entry.
  # Location input is the experimental site to use later in joins
  f_to_c <- 5/9
  clean_prism <- raw_prism %>%
    mutate(Year = year(date),
           doy = yday(date),
           maxt_c = (maxt-32)*f_to_c,
           mint_c = (mint-32)*f_to_c,
           prcp_cm = prcp*2.54,
           Location = rep(location ,nrow(raw_prism))) %>%
  dplyr::select(Year, date, doy, maxt_c, mint_c, prcp_cm, Location) %>%
    filter(Year > 1980)
  return(clean_prism)
}

# Now you can get your metric prism data for each location by just referencing the file path
musgrave_prism <- PrismRaw2Clean(csvpath = "R_data/Weather_data/PRISM/Raw_prism_files/musgrave_prism.csv", location = "Musgrave")
geneva_prism <- PrismRaw2Clean(csvpath = "R_data/Weather_data/PRISM/Raw_prism_files/geneva_prism.csv", location = "Geneva")
arlington_prism <- PrismRaw2Clean(csvpath = "R_data/Weather_data/PRISM/Raw_prism_files/arlington_prism.csv", location = "AARS")
psu_prism <- PrismRaw2Clean(csvpath = "R_data/Weather_data/PRISM/Raw_prism_files/psu_prism.csv", location = "Rock Springs")
```

```{r Precipitation Frequency Analysis}

prism_list <- list(musgrave_prism, arlington_prism, geneva_prism, psu_prism)

process_weather_data <- function(weather_data) {
  
  # Define the three 30-year periods (adjust according to your data range)
  periods <- list(
    "early" <- 1981:1995,
  "middle" <- 1996:2010,
  "late" <- 2011:2024
  )
  
  # Function to calculate annual maxima for each period
  calculate_annual_max <- function(year_range) {
    weather_data %>%
      filter(Year %in% year_range) %>%
      group_by(Year) %>%
      summarize(max_precip = max(prcp_cm, na.rm = TRUE))
  }
  
  # Get annual maxima for each period
  annual_max_list <- lapply(periods, calculate_annual_max)
  
  # Function to fit lognormal distribution and calculate return levels
  fit_lognormal <- function(annual_max) {
    annual_max <- annual_max$max_precip[annual_max$max_precip > 0]  # Remove zero or negative values
    
    fit <- fitdist(annual_max, "lnorm")  # Fit lognormal distribution
    return(fit)
  }
  
  # Fit lognormal distribution to each period
  lognormal_fits <- lapply(annual_max_list, fit_lognormal)
  
  # Calculate return levels
  return_periods <- c(5, 10, 15, 20, 25, 30)
  
  calculate_return_levels <- function(fit) {
    # Extract parameters
    meanlog <- fit$estimate["meanlog"]
    sdlog <- fit$estimate["sdlog"]
    
    # Calculate return levels
    return_levels <- qlnorm(1 - (1 / return_periods), meanlog = meanlog, sdlog = sdlog)
    
    # Return levels as a data frame
    data.frame(
      Return_Period = return_periods,
      Return_Level = return_levels
    )
  }
  
  # Calculate return levels for each period
  return_levels_list <- lapply(lognormal_fits, calculate_return_levels)
  
  # Combine results into a single data frame for all periods
  return_levels_df <- bind_rows(
    data.frame(Period = "early", return_levels_list[[1]]),
    data.frame(Period = "middle", return_levels_list[[2]]),
    data.frame(Period = "late", return_levels_list[[3]])
  )
  
  return(return_levels_df)
}

return_times_sites <- lapply(prism_list, process_weather_data)

combined_results <- bind_rows(
  lapply(seq_along(return_times_sites), function(i) {
    data.frame(Data_Set = paste("Data", i), return_times_sites[[i]])
  })
)

combined_results <- combined_results %>%
  mutate(site = case_match(Data_Set, 
                           "Data 1" ~ "Musgrave",
                           "Data 2" ~ "AARS",
                           "Data 3" ~ "Geneva",
                           "Data 4" ~ "Rock Springs"))

# Plot rainfall distributions of return times for each site and color dots by the time period
combined_results %>%
  ggplot(aes(x = Return_Period, y = Return_Level, color = Period)) +
  geom_point() +
  geom_line() +
  facet_wrap(~site) + 
  theme_bw() +
  labs(y = "Daily Precipitation Amount (cm)",x = "Return period (Years)", legend = "Time period" )

```


```{r Read in soybean agronomic data}
# Read in all soybean data from raw file
soy_rye_raw <- read_csv(here("R_data/Soy_rye_data/NY_WI_NTvT_master.csv"))

# Select only rows that can be useful for further analysis
# Also editing experiment name for 2018 so I can model it differently
soy_rye_useful <- soy_rye_raw %>%
  mutate(ExperimentName = case_when(Field == "RN30" ~ "White Mold",
                                    Field == "RN20" ~ "White Mold",
                                    .default = ExperimentName)) %>%
  select(Year, State, Location, ExperimentName, Rep, SoySeedingDate, Till_NT,
         RyeBiomass_kgha, SoyStand_ha, Per_Emergence, WeedBiomass_kgha, Yield_Mgha, Yield_buac, ExperimentalTrt) %>%
  filter(ExperimentName != "CIG NT Soybeans") %>%
  mutate(SoySeedingDate = as.Date(SoySeedingDate, format = "%m/%d/%y")) %>%
  filter(SoySeedingDate != "2019-10-06") %>%
  mutate(ExperimentName = case_when(SoySeedingDate == "2018-06-06"~ "No-Till Soy Split Plot",
                                    .default = ExperimentName),
         Location = case_when(State == "PA" ~ "Rock Springs", .default = Location))


# Count number of no-till and tilled plots for each experiment and year of the experiment
nt_t_counts <- soy_rye_useful %>%
  group_by(ExperimentName, Year) %>%
  mutate(nt_count = sum(Till_NT == "NT")) %>%
  mutate(till_count = sum(Till_NT == "Till")) %>%
  distinct(ExperimentName, till_count, nt_count)
```

```{r Functions to get meta analysis independent variables}
#### These functions are for getting independent variables for later use in meta analysis, such as precip before and after planting, GDD, etc. These are variables that are climatically based and vary by site year
planting_precip <- function(met_df, mgmt_df, station_site){

  # met_df: Dataframe that includes precipitation data, year, and date at least to calculate precipitation around planting.
  # mgmt_df: Dataframe that includes soybean management and agronomic information, including planting date
  # station_site: String input for the name of experimental site (Musgrave, Arlington, Geneva, PSU, etc.)

# Find planting dates for the location for each experiment and year
  planting_dates <- mgmt_df %>%
    group_by(Year, Location) %>%
    slice_min(SoySeedingDate) %>%
    ungroup() %>%
    filter(Location == station_site) %>%
    select(ExperimentName, Year, State, Location, SoySeedingDate) %>%
    distinct()
  
  exp_df <- left_join(met_df, planting_dates, by = c("Year","Location")) %>%
    mutate(plant_doy = yday(SoySeedingDate))
#### Calculates the precipitation +/- 15 days around planting period for each planting date at the site 
  plant_prcp <- exp_df %>%
    filter(!is.na(SoySeedingDate) & doy >= plant_doy - 15 & doy<= plant_doy + 15) %>%
    group_by(ExperimentName, Year, State, Location, SoySeedingDate) %>%
    summarise(plant_prcp = sum(prcp_cm, na.rm = T), .groups = "drop")
  return(plant_prcp)
}

#More functions to come depending on variables that we investigate, but format will be similar regardless of variable
musgrave_plant_prcp <- planting_precip(met_df = musgrave_prism, mgmt_df = soy_rye_useful, station_site = "Musgrave")
wi_plant_prcp <- planting_precip(met_df = arlington_prism, mgmt_df = soy_rye_useful, station_site = "AARS")
geneva_plant_prcp <- planting_precip(met_df = geneva_prism, mgmt_df = soy_rye_useful, station_site = "Geneva")
psu_plant_prcp <- planting_precip(met_df = psu_prism,
                                  mgmt_df = soy_rye_useful, 
                                  station_site = "Rock Springs")

all_plant_prcp <- rbind(musgrave_plant_prcp, wi_plant_prcp, geneva_plant_prcp, psu_plant_prcp)
comb_soy_prcp <- left_join(soy_rye_useful, all_plant_prcp, by = c("Year", "State", "Location", "ExperimentName"))
```

```{r Function to get each experiments yield data set up for meta analysis}

get_model_inputs <- function(soy_data, experiment_name){
  # This function gets the data for each individual experiment to be used for creating each model
  model_input <- soy_data %>%
    filter(ExperimentName %in% experiment_name) %>%
    drop_na(Year) %>%
    mutate(Year = as.factor(Year)) %>%
    mutate(block = Rep)
}  
 
experiment_names <- unique(comb_soy_prcp$ExperimentName)

# For loop below used to easily extract each dataframe for my experiments
input_list <- list()
for(i in 1:length(experiment_names)){
  input_list[[i]] <- get_model_inputs(comb_soy_prcp, experiment_names[i])
  clean_name <- gsub("[^a-zA-Z0-9]", "_", tolower(experiment_names[i]))
  df_name <- paste0(clean_name, "_inputs")
  assign(df_name, input_list[[i]])
}

```

```{r Creating Mixed Models}
# Models will be created outside of functions or loops to ensure good model fitting techniques

# Barns
barns_mod <- lmer(log(Yield_Mgha) ~ Till_NT*Year + (1|Year:block), data = barns_inputs)
barns_glm <- glmmTMB(Yield_Mgha ~ Till_NT*Year + (1|Year:block),
                        family = tweedie(link = "log"),
                        data = barns_inputs)
summary(barns_glm)
simulateResiduals(barns_glm, plot = TRUE)

# MOSS 
moss_mod <- lmer(log(Yield_Mgha) ~ Till_NT*Year + (1|Year:block), data = moss_inputs)
summary(moss_mod)
anova(moss_mod, ddf = "Kenward-Roger")
check_model(moss_mod)
```

 exp_mod <- if(length(unique(soy_data$Year[soy_data$ExperimentName %in% experiment_name])) > 1){
        lmer(log(Yield_Mgha) ~ as.factor(Till_NT)*as.factor(Year) + (1|Year:block) + (1:Year), data = model_input)
  } else{
    lmer(log(Yield_Mgha) ~ as.factor(Till_NT)*Year + (1|block), data = model_input)
  }
```{r Getting model contrasts}
get_model_contrasts <- function(exp_model, experiment_name){
  # Function takes input of a model and experiment name to get out the contrasts from emmeans
  mod_emm <- emmeans(exp_mod,pairwise ~Till_NT|Year)
 
  contrast_df <- as.data.frame(mod_emm$contrasts) %>%
    select(Year, estimate, SE) %>%
    mutate(ExperimentName = experiment_name) %>%
    rename(yield_gap = estimate)
  
  return(contrast_df)
}  

# Run function for all of my data
emm_list <- list()
for(i in 1:length(experiment_names)){
  emm_list[[i]] <- get_model_contrasts(model_list[i],exp)
}
```



barns_emm <- get_soy_model(comb_soy_prcp, experiment_name = "BARNS")
moss_emm <- get_soy_model(comb_soy_prcp, experiment_name = "MOSS")
ocs_emm <- get_soy_model(comb_soy_prcp, experiment_name = "OCS")
coulter_emm <- get_soy_model(comb_soy_prcp, experiment_name = "Coulter Study")
sf_emm <- get_soy_model(comb_soy_prcp, experiment_name = "Starter Fertilizer Study")
sd_emm <- get_soy_model(comb_soy_prcp, experiment_name = "Seeding Depth Study")
mowtivation_emm <- get_soy_model(comb_soy_prcp, experiment_name = "Mowtivation")
wi_nt_soy_emm <- get_soy_model(comb_soy_prcp, experiment_name = "No-Till Soybeans")
wi_ccbrt_emm <- get_soy_model(comb_soy_prcp, experiment_name = "CCBRT Soybean")
rose2_emm <- get_soy_model(comb_soy_prcp, experiment_name = "ROSE2")
rose3_emm <- get_soy_model(comb_soy_prcp, experiment_name = "ROSE3")

# Separate model for WI 2018 split plot setup
wi_2018 <- comb_soy_prcp %>%
  filter(ExperimentName == "No-Till Soy Split Plot") %>%
  separate_wider_delim(ExperimentalTrt, delim = " ", names = c("main_plot","split_plot"))
wi_2018_mod <- lmer(log(Yield_Mgha) ~ as.factor(Till_NT)*Year + (1|Rep:main_plot) + (1:Rep), data = wi_2018)
hist(residuals(wi_2018_mod))
plot(residuals(wi_2018_mod)~predict(wi_2018_mod))
wi_2018_emm <- emmeans(wi_2018_mod, pairwise ~ Till_NT|Year)
wi_2018_contrast <- as.data.frame(wi_2018_emm$contrasts) %>%
    select(Year, estimate, SE) %>%
    mutate(ExperimentName = "No-Till Soy Split Plot") %>%
    rename(yield_gap = estimate)


# Combine all of the contrasts into one dataframe
total_contrasts <- rbind(barns_emm[[1]], moss_emm[[1]], ocs_emm[[1]],
                         mowtivation_emm[[1]], coulter_emm[[1]], sf_emm[[1]], sd_emm[[1]],
                         wi_nt_soy_emm[[1]], wi_ccbrt_emm[[1]], wi_2018_contrast,
                         rose2_emm[[1]], rose3_emm[[1]])

```

```{r Meta analysis work}
# Create inpput for meta analysis
meta_input <- left_join(total_contrasts, comb_soy_prcp) %>%
  select(Year, State, Location, ExperimentName, SoySeedingDate.y, yield_gap, SE, plant_prcp) %>%
  rename("SoySeedingDate" = SoySeedingDate.y) %>%
  mutate(site_year = paste0(State, "-", Year)) %>%
  distinct()

meta_input$SoySeedingDate[meta_input$ExperimentName == "Starter Fertilizer Study"] = as.Date("2021-06-04", format = "%Y-%m-%d")
meta_input$plant_prcp[meta_input$ExperimentName == "Starter Fertilizer Study"] = 7.44


# Run the meta analysis
metagen_analysis <- metagen(data = meta_input, TE = yield_gap, seTE = SE, studlab = site_year)

# Save the results into this dataframe
metagen_result <- data.frame(
  effect_size = metagen_analysis$TE,  # Treatment effect
  ci_lower = metagen_analysis$lower,  # Lower bound of confidence interval
  ci_upper = metagen_analysis$upper,  # Upper bound of confidence interval
  i2 = metagen_analysis$I2            # I² for heterogeneity
)

# Combine with identifying data and rainfall data
metagen_result_comb <- cbind(metagen_result, meta_input)

# Transform the outputs to a linear scale from the log scale
metagen_result_final_1 <- metagen_result_comb %>%
  mutate(yield_ratio = exp(yield_gap),
         lowerci_trans = exp(ci_lower),
         upperci_trans = exp(ci_upper)) %>%
  filter(site_year != "PA-2016" & site_year != "PA-2023" & site_year != "PA-2022")

# Plot the output of the meta analysis
ratio_yg_meta_1 <- metagen_result_final_1 %>%
  ggplot(aes(x = plant_prcp, y = yield_ratio))+
  geom_point(aes(color = State), size = 2) +
  geom_errorbar(aes(ymin = lowerci_trans, ymax = upperci_trans, colour = State), width = 0.2) + # Error bars for CI
  #geom_text(aes(label  = site_year)) +
  labs(x = "Precipitation accumulation +/- 15 days of planting (cm)",
       y = "Ratio of no-till yield to tilled yield",
       color = "") +
  stat_poly_eq(formula = y~x, use_label(c("R2", "eq")), size = 5)+
  theme_pubr() +
  scale_color_manual(values = c("black", "slateblue", "red")) +
  geom_smooth(method = "lm",formula = y ~ x,  se = FALSE, color = "gray", 
              linetype = "dashed")  +
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16))
```
